---
title: "Untitled"
output: html_document
date: "2025-08-06"
---

### This script is used for constructing figures 4B, S7 & S9B. 

## Load pacakges

```{r}

library(ggplot2)

library(dplyr)

library(readr)

library(mgcv)

```

Load datasets

```{r}

amazondf <- read.csv('C:\\Users\\haozh\\Desktop\\buttress_root\\buttress_plot_pixel_aggregate_20251219.csv');


```



```{r}
amazondf <- amazondf %>%
   dplyr::select(matches(
    'CHELSA|SG|Canopy|random|pixel|Earth|Depth|Fire|Land|Nitrogen|Olsen|Organic|PET|WWF|World|fraction|buttress|wind|Distance|Aridity'
  ))%>%
  mutate(longitude = pixellon,
         latitude = pixellat)


amazondf <- amazondf[complete.cases(amazondf),]
```

Tropical points

```{r}
tropdf <- amazondf %>%
  #filter(fraction==0)%>%
  filter(WWF_Biome %in% c(1,2,3,7))%>%
  mutate(majorbiome = 'Tropical')


nrow(tropdf)

```

Temperate points

```{r}
set.seed(0)

tempdf <- amazondf %>%
  filter(WWF_Biome %in% c(4,5,8,12))%>%
  sample_n(round(nrow(tropdf)*22/50))%>%
  mutate(majorbiome = 'Temperate')

nrow(tempdf)

```
Boreal points 

```{r}
set.seed(0)

borealdf <- amazondf %>%
  filter(WWF_Biome %in% c(6,10,11))%>%
  #sample_n(round(nrow(tropdf)*21/50))%>%
  mutate(majorbiome = 'Boreal')


```
Other points

```{r}
set.seed(0)
drydf <- amazondf %>%
  filter(WWF_Biome %in% c(9,13))%>%
  sample_n(round(nrow(tropdf)*7/50))%>%
  mutate(majorbiome = 'Arid_and_other')

nrow(drydf)

```
Combine them together

```{R}
alldf <- rbind(tropdf, tempdf, borealdf, drydf)

alldf 

```

```{r}

regressionmatrix <- alldf %>%
  #filter(WWF_Biome %in% c(1))%>% ## This should be used if only test tropical only dataset.
  dplyr::select(c(
    CHELSA_Annual_Mean_Temperature,
    wind_gust_max,
    CHELSA_Annual_Precipitation,
    CHELSA_swb,
    DistanceToForestEdge,
    Nitrogen,
    OlsenP,
    SG_Soil_pH_H2O_0_100cm,
    SG_Silt_Content_0_100cm,
    SG_Absolute_depth_to_bedrock,
    meanfraction
  ))

regressionmatrix <- regressionmatrix[complete.cases(regressionmatrix),]

```

The following chuncks can be used to select the best distribution facmily and link functions. 

```{r}
themodel0 <- glm(-log(2-meanfraction)~ scale(Nitrogen)+scale(SG_Silt_Content_0_100cm) + scale(OlsenP) + scale(CHELSA_Annual_Precipitation) +scale(SG_Soil_pH_H2O_0_100cm)+scale(CHELSA_Annual_Mean_Temperature)  + scale(wind_gust_max) + scale(SG_Absolute_depth_to_bedrock) + scale(CHELSA_swb) +  scale(wind_gust_max*SG_Soil_pH_H2O_0_100cm), data = regressionmatrix)



themodel1 <- glm(meanfraction~ scale(Nitrogen)+scale(SG_Silt_Content_0_100cm) + scale(OlsenP) + scale(CHELSA_Annual_Precipitation) +scale(SG_Soil_pH_H2O_0_100cm)+scale(CHELSA_Annual_Mean_Temperature)  + scale(wind_gust_max) + scale(SG_Absolute_depth_to_bedrock) + scale(CHELSA_swb) + scale(wind_gust_max*SG_Soil_pH_H2O_0_100cm), data = regressionmatrix, family = binomial(link = 'cloglog'))

themodel2 <- glm(meanfraction~ scale(Nitrogen)+scale(SG_Silt_Content_0_100cm) + scale(OlsenP) + scale(CHELSA_Annual_Precipitation) +scale(SG_Soil_pH_H2O_0_100cm)+scale(CHELSA_Annual_Mean_Temperature)  + scale(wind_gust_max) + scale(SG_Absolute_depth_to_bedrock) + scale(CHELSA_swb) , data = regressionmatrix,, family = binomial(link = 'probit'))

themodel3 <- glm(meanfraction~ scale(Nitrogen)+scale(SG_Silt_Content_0_100cm) + scale(OlsenP) + scale(CHELSA_Annual_Precipitation) +scale(SG_Soil_pH_H2O_0_100cm)+scale(CHELSA_Annual_Mean_Temperature)  + scale(wind_gust_max) + scale(SG_Absolute_depth_to_bedrock) + scale(CHELSA_swb) , data = regressionmatrix,, family = binomial(link = 'logit'))

gammodel <- gam(meanfraction~ scale(Nitrogen)+scale(SG_Silt_Content_0_100cm) + scale(OlsenP) + scale(CHELSA_Annual_Precipitation) +scale(SG_Soil_pH_H2O_0_100cm)+scale(CHELSA_Annual_Mean_Temperature)  + scale(wind_gust_max) + scale(SG_Absolute_depth_to_bedrock) + scale(CHELSA_swb)  , data = regressionmatrix,)

AIC(themodel0, themodel1, themodel2, themodel3, gammodel)

BIC(themodel0, themodel1, themodel2, themodel3, gammodel)

```

Loop arounnd random seeds and subset 10% of the gloal subsampled dataset and apply the glm model. 

```{r}
for (i in 1:100) {
  
  set.seed(i)
  subdf <- regressionmatrix%>%
    sample_n(round(0.1*nrow(regressionmatrix)))
  
  themodel1 <- glm(meanfraction~ scale(Nitrogen)+scale(SG_Silt_Content_0_100cm) + scale(OlsenP) + scale(CHELSA_Annual_Precipitation) +scale(SG_Soil_pH_H2O_0_100cm)+scale(CHELSA_Annual_Mean_Temperature)  +scale(wind_gust_max) + scale(SG_Absolute_depth_to_bedrock) + scale(DistanceToForestEdge) 
                     #scale(SG_Absolute_depth_to_bedrock*wind_gust_max) + 
                     #scale(SG_Soil_pH_H2O_0_100cm*OlsenP)+
                     #scale(SG_Soil_pH_H2O_0_100cm*Nitrogen)
                     , data = subdf, family = binomial(link = 'cloglog'))

  
  ss <- summary(themodel1)



modelsummary <- themodel1$coefficients


ssd <- ss$coefficients


ssd <- ssd[2:nrow(ssd),]



ssd <- as.data.frame(ssd)


items <- rownames(ssd)

row.names(ssd) <- NULL

ssd <- cbind(items, ssd)
names(ssd) <- c('items', 'Estimate', 'stderr', 'zvalue', 'pvalue')

ssd <- ssd %>%
  mutate(seed = i)
  
write_csv(ssd, paste0('C:\\Users\\haozh\\Desktop\\buttress_root\\lmcoef\\lmcoefs_subsampleall_seed',i,'.csv'))

}


```


```{r}

filelist <- list.files(path = 'C:\\Users\\haozh\\Desktop\\buttress_root\\lmcoef', pattern = "\\.csv", full.names = TRUE)

allsubs <-do.call(bind_rows,
          lapply(filelist, read.csv))

```

### Construct the figure....
```{r}

allsubs$items <- factor(allsubs$items, levels = c(
  
  
  
  #'scale(SG_Clay_Content_0_100cm)',
  'scale(OlsenP)',
  'scale(Nitrogen)',
  #'scale(SG_Soil_pH_H2O_0_100cm * Nitrogen)',
  #'scale(SG_Soil_pH_H2O_0_100cm * OlsenP)',
  #'scale(SG_Absolute_depth_to_bedrock * wind_gust_max)',
  
  
  "scale(SG_Absolute_depth_to_bedrock)",
  "scale(DistanceToForestEdge)",
  'scale(SG_Silt_Content_0_100cm)',
  'scale(SG_Soil_pH_H2O_0_100cm)',
  'scale(wind_gust_max)',
  'scale(CHELSA_Annual_Precipitation)',
  #'scale(CHELSA_npp)',
  'scale(CHELSA_Annual_Mean_Temperature)'
))

allsubs %>%
  ggplot(., aes(x = items, y = Estimate))+
  stat_summary(fun.data = 'mean_cl_boot', size = 1, linewidth = 1)+
  coord_flip()+
  scale_color_gradient(low = "#2166AC", high = "#B2182B")+
  theme_classic()+
  geom_abline(slope = 0, intercept = 0,linetype = 'dashed')+
  labs(x = 'Variables', y = 'Effect')+
  theme(legend.position = 'none',  axis.title.x = element_text(color = '#000000', size = 20), , axis.text.x = element_text(size = 20, colour = 'black'), , axis.title.y = element_blank())



```



```{r}
formlist <- c(
  'meanfraction~scale(CHELSA_Annual_Mean_Temperature)',
  #'meanfraction~scale(CHELSA_npp)',
  'meanfraction~scale(CHELSA_Annual_Precipitation)',
  'meanfraction ~ scale(CHELSA_Annual_Mean_Temperature * CHELSA_Annual_Precipitation)',
  'meanfraction~scale(wind_gust_max)',
  'meanfraction~scale(SG_Soil_pH_H2O_0_100cm)',
  'meanfraction~scale(SG_Silt_Content_0_100cm)',
  'meanfraction~scale(SG_Absolute_depth_to_bedrock)',
  'meanfraction~scale(SG_Absolute_depth_to_bedrock * wind_gust_max)',
  'meanfraction~scale(SG_Soil_pH_H2O_0_100cm * Nitrogen)',
  'meanfraction~scale(SG_Soil_pH_H2O_0_100cm * OlsenP)',
  #'meanfraction~scale(CHELSA_swb)',
  'meanfraction~scale(DistanceToForestEdge)',
  'meanfraction~scale(OlsenP)',
  'meanfraction~scale(Nitrogen)'
)



```

### Univariate models
```{r, warning=FALSE}

for(j in formlist){

for (i in 101:200) {
  
  set.seed(i)
  subdf <- regressionmatrix%>%
    sample_n(round(0.1*nrow(regressionmatrix)))
  
  themodel1 <- glm(j , data = subdf, family = binomial(link = 'cloglog'))

  
  ss <- summary(themodel1)

itemname = substr(j,14,nchar(j))
itemname = gsub("\\*", "x", itemname)

modelsummary <- themodel1$coefficients


ssd <- ss$coefficients
ssd <- as.data.frame(ssd)

ssd <- ssd[2:nrow(ssd),]



ssd <- as.data.frame(ssd)


#ssd <- t(ssd)

items <- j

row.names(ssd) <- NULL

ssd <- cbind(items, ssd)
names(ssd) <- c('items','Estimate', 'stderr', 'zvalue', 'pvalue')

ssd <- ssd %>%
  mutate(seed = i)
  
write_csv(ssd, paste0('C:\\Users\\haozh\\Desktop\\buttress_root\\lm_uni\\unilmcoefs_subsampleall_',itemname,'_seed',i,'.csv'))

}
}


```


```{r}

filelist <- list.files(path = 'C:\\Users\\haozh\\Desktop\\buttress_root\\lm_uni', pattern = "\\.csv", full.names = TRUE)

allsubs <-do.call(bind_rows,
          lapply(filelist, read.csv))

```

```{r}

allsubs$items <- factor(allsubs$items, levels = c(

  
  
  'meanfraction~scale(OlsenP)',
  'meanfraction~scale(Nitrogen)',
  'meanfraction~scale(SG_Absolute_depth_to_bedrock)',
  'meanfraction~scale(DistanceToForestEdge)',
  
  'meanfraction~scale(SG_Soil_pH_H2O_0_100cm * OlsenP)',
  'meanfraction~scale(SG_Soil_pH_H2O_0_100cm * Nitrogen)',
  'meanfraction~scale(SG_Absolute_depth_to_bedrock * wind_gust_max)',
  #'meanfraction~scale(CHELSA_swb)',
  
  'meanfraction~scale(SG_Silt_Content_0_100cm)',
  'meanfraction~scale(SG_Soil_pH_H2O_0_100cm)',
  'meanfraction~scale(wind_gust_max)',
  'meanfraction ~ scale(CHELSA_Annual_Mean_Temperature * CHELSA_Annual_Precipitation)',
  'meanfraction~scale(CHELSA_Annual_Precipitation)',
  'meanfraction~scale(CHELSA_Annual_Mean_Temperature)'
))

allsubs %>%
  ggplot(., aes(x = items, y = Estimate))+
  stat_summary(fun.data = 'mean_cl_boot', size = 1, linewidth = 1)+
  coord_flip()+
  scale_color_gradient(low = "#2166AC", high = "#B2182B")+
  theme_classic()+
  geom_abline(slope = 0, intercept = 0,linetype = 'dashed')+
  labs(x = 'Variables', y = 'Effect')+
  theme(legend.position = 'none',  axis.title.x = element_text(color = '#000000', size = 20), , axis.text.x = element_text(size = 20, colour = 'black'))



```


