---
title: "Untitled"
output: html_document
date: "2024-07-26"
---

### This script is used for understanding environmental drivers of buttress root both globally and in tropical rainforests. 

### This script constructs Figure 4A and Figures S6 & S10.


Load packages

```{r}
library(data.table)
library(h2o)
library(tictoc)
library(foreach)
library(doParallel)
library(outliers)
library(EnvStats)
library(zoo)
library(mgcv)
library(nlme)
library(HH)
library(ggplot2)
library(dplyr)
library(readr)

library(maps)

worlddata <- map_data('world')


```

Load dataset
```{r}

amazondf <- read.csv('buttress_plot_pixel_aggregate_20251219.csv');

```


```{r}
amazondf <- amazondf %>%
   dplyr::select(matches(
    'CHELSA|SG|Canopy|random|pixel|Earth|Depth|Fire|Land|Nitrogen|Olsen|Organic|PET|WWF|World|fraction|buttress|wind|Distance|Aridity'
  ))%>%
  mutate(longitude = pixellon,
         latitude = pixellat)


amazondf <- amazondf[complete.cases(amazondf),]
```

```{r}
amazondf <- amazondf %>%
  filter(!is.na(CHELSA_Annual_Mean_Temperature))%>%
  filter(!is.na(CHELSA_Annual_Precipitation))%>%
  filter(!is.na(wind_gust_max))%>%
  filter(!is.na(Nitrogen))%>%
  filter(!is.na(OlsenP))%>%
  filter(!is.na(SG_Soil_pH_H2O_0_100cm))%>%
  filter(!is.na(SG_Silt_Content_0_100cm))%>%
  filter(!is.na(SG_Absolute_depth_to_bedrock))%>%
  filter(!is.na(DistanceToForestEdge))


```

Tropical subset

```{r}
tropdf <- amazondf %>%
  #filter(fraction==0)%>%
  filter(WWF_Biome %in% c(1,2,3,7))%>%
  mutate(majorbiome = 'Tropical')



nrow(tropdf)

```


Temperate subset


```{r}
set.seed(0)

tempdf <- amazondf %>%
  filter(WWF_Biome %in% c(4,5,8,12))%>%
  sample_n(round(nrow(tropdf)*22/50))%>%
  mutate(majorbiome = 'Temperate')

nrow(tempdf)

```

Boreal subset
```{r}
set.seed(0)

borealdf <- amazondf %>%
  filter(WWF_Biome %in% c(6,10,11))%>%
  #sample_n(round(nrow(tropdf)*21/50))%>%
  mutate(majorbiome = 'Boreal')


```
Other area subset

```{r}
set.seed(0)
drydf <- amazondf %>%
  filter(WWF_Biome %in% c(9,13))%>%
  sample_n(round(nrow(tropdf)*7/50))%>%
  mutate(majorbiome = 'Arid_and_other')

nrow(drydf)
```



```{R}
alldf <- rbind(tropdf, tempdf, borealdf, drydf)

alldf 

```

```{r}

regressionmatrix <- alldf %>%
  #filter(PlotObservationID %in% tallbuttress$PlotObservationID)%>%
  #filter( CHELSA_gdd10>40000 & CHELSA_swb>-500)%>%
  #filter(meanfraction>0)%>%
  #filter(WWF_Biome %in% c(1))%>%  ## This should be added if only tropical areas are being tested. 
  dplyr::select(c(
    CHELSA_Annual_Mean_Temperature,
    #CHELSA_Min_Temperature_of_Coldest_Month,
    #CHELSA_Precipitation_of_Driest_Month,
    #CHELSA_Wind_max,
    wind_gust_max,
    CHELSA_Annual_Precipitation,
    #CHELSA_Temperature_Annual_Range,
    #CHELSA_Mean_Temperature_of_Wettest_Quarter,
    #CHELSA_Annual_Precipitation,
    #CHELSA_swb,
    #CHELSA_npp,
    #CHELSA_gdd10,
    #CHELSA_Precipitation_Seasonality,
    #CHELSA_Temperature_Seasonality,
    #CHELSA_Temperature_Annual_Range,
    Nitrogen,
    OlsenP,
    #np_limitation,
    #cnRatio,
    #NDVI,
    SG_Soil_pH_H2O_0_100cm,
    #EarthEnvTopoMed_Roughness,
    #EarthEnvTopoMed_Elevation,
    #CanopyHeight,
    #PresentTreeCover,
    #NDVI,
    #EarthEnvTopoMed_Slope,
    #PresentTreeCover,
    #CanopyHeight,
    SG_Silt_Content_0_100cm,
    SG_Absolute_depth_to_bedrock,
    DistanceToForestEdge,
    #SG_Sand_Content_0_100cm,
    #SG_Clay_Content_0_100cm,
    #PET,
    #Depth_to_Water_Table,
    #SG_Clay_Content_0_100cm,
    #SG_Silt_Content_0_100cm,
    #WorldClim2_WindSpeed_AnnualMean,
    #SG_Soil_pH_H2O_0_100cm,
    meanfraction
  ))

regressionmatrix <- regressionmatrix[complete.cases(regressionmatrix),]

```

```{r}

regressionmatrix %>%
  filter(meanfraction>0)%>%
  nrow()

```

Test colinearity 

```{r}
HH::vif(regressionmatrix[,-ncol(regressionmatrix)])
```

```{R}
cor.matrix <- cor(regressionmatrix[,1:(ncol(regressionmatrix)-1)], use = 'complete.obs', method = 'pearson')

cor.dist <- abs(as.dist(cor.matrix))

cor.cluster <- hclust(1-cor.dist)

plot(cor.cluster)

```



Initiate environment for random forest models

```{R}

localH2O <- h2o.init(nthreads = 12, max_mem_size = '16g', ignore_config = TRUE)

regmatrixh2o <- as.h2o(regressionmatrix)

rf.params <- list(ntrees = 250,
                  mtries = c(1:8),
                  min_rows = c(1:20))

search.criteria <- list(strategy = 'RandomDiscrete', max_models = 160, seed = 0, max_runtime_secs = 1200)

```



```{R}
rf.grid <- h2o.grid(algorithm = 'randomForest',
                    y = 'meanfraction',
                    grid_id = 'rf.grid',
                    training_frame = regmatrixh2o,
                    seed = 0,
                    hyper_params = rf.params,
                    sample_rate = 0.632,
                    nfolds = 10,
                    fold_assignment = 'AUTO',
                    keep_cross_validation_predictions = TRUE,
                    keep_cross_validation_fold_assignment = TRUE,
                    search_criteria = search.criteria)


```

```{R}
rf.grid.perf = h2o.getGrid(grid_id = 'rf.grid',
                           sort_by = 'R2',
                           decreasing = TRUE)

head(rf.grid.perf@summary_table,10)

```


```{R}

for(i in 1:10){
  rfmodel <- h2o.getModel(rf.grid.perf@model_ids[[i]])
  
  p1<- h2o.varimp_plot(rfmodel)
  
  print(p1)
  
  varip <- h2o.varimp(rfmodel)
  
  varip
  
  library(readr)
  
  
  
}

```

Get the shap value of the best model. 
```{R}
rf_model <- h2o.getModel(rf.grid.perf@model_ids[[1]])

shapsummary <- h2o.shap_summary_plot(rf_model, regmatrixh2o, sample_size = nrow(regressionmatrix))

shapsummary


```


```{R}
shapdata <- shapsummary$data

```


```{R}
h2o.shutdown(prompt = FALSE)

```


```{R}
shapdata$original_value <- as.numeric(shapdata$original_value)

shapdata %>%
  dplyr::group_by(feature) %>%
  dplyr::summarise(count = n()) %>%
  ungroup()


plots = shapdata %>%
  group_by(feature) %>%
  group_map(~ggplot(data = ., aes(x = original_value, y = contribution))+
              geom_point(alpha = 0.5)+
              geom_smooth(method='loess')+
              geom_hline(yintercept = 0))

```
```{r}

meanshap <- shapdata %>%
  group_by(feature)%>%
  dplyr::summarise(
    meanabs = mean(abs(contribution)),
    medianabs = median(abs(contribution)),
    lowerlimit = quantile(contribution, 0.025),
    upperlimit = quantile(contribution, 0.975),
    median = median(contribution)
  )%>%
  ungroup()
 

```



```{r}
shapdata$feature <- factor(shapdata$feature, levels = c(
                                                        
                                                        
                                                        
                                                        
                                                        
                                                        'OlsenP',
                                                        
                                                        'Nitrogen', 
                                                        'SG_Absolute_depth_to_bedrock',
                                                        'DistanceToForestEdge',
                                                       'SG_Silt_Content_0_100cm',
                                                       'SG_Soil_pH_H2O_0_100cm',
                                                       
                                                        'wind_gust_max',
                                                        'CHELSA_Annual_Precipitation',
                                                        #'CHELSA_gdd10',
                                                        #'CHELSA_npp',
                                                        'CHELSA_Annual_Mean_Temperature'
                                                        
                                                        
                                                        #'CHELSA_Min_Temperature_of_Coldest_Month'
                                                        ))


shapdata %>%
  mutate(abscont = abs(contribution))%>%
  ggplot(., aes(x = feature, y = abscont))+
  stat_summary(fun = mean, geom = 'bar')+
  stat_summary(fun.data = 'mean_cl_boot', geom = 'errorbar',size = 1, width = 0.6)+
  labs(x = 'Variables', y = 'Mean absolute SHAP value')+
  coord_flip()+
  theme_classic()+
  theme(axis.title = element_text(color = '#000000', size = 20), axis.text.x = element_text(color = '#000000', size = 20))


```

Construct SHAP plots

```{R}
ppx<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'wind_gust_max') %>%
  group_map(~ggplot(data = ., aes(x = original_value, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.4)+
              labs(x = 'Wind gust (m/s)', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              coord_cartesian(ylim = c(-0.07,0.15))+
              #scale_x_continuous(trans = 'log10')+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))

ppy<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'DistanceToForestEdge') %>%
  group_map(~ggplot(data = ., aes(x = original_value, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.4)+
              labs(x = 'Distance to forest edge (meters)', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              coord_cartesian(ylim = c(-0.07,0.15))+
              scale_x_continuous(trans = 'log1p')+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))


pp0<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'CHELSA_npp') %>%
  group_map(~ggplot(data = ., aes(x = original_value/10, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.4)+
              labs(x = 'Potential NPP', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              #scale_x_continuous(trans = 'log10')+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))


pp1<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'CHELSA_gdd10') %>%
  group_map(~ggplot(data = ., aes(x = original_value/10, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.4)+
              labs(x = 'Growing degree day length (\u00B0C)', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              #scale_x_continuous(trans = 'log10')+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))




pp<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'CHELSA_Annual_Mean_Temperature') %>%
  group_map(~ggplot(data = ., aes(x = original_value/10, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.4)+
              labs(x = 'Annual mean temperature (\u00B0C)', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              coord_cartesian(ylim = c(-0.07,0.15))+
              #scale_x_continuous(trans = 'log10')+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))

p<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'CHELSA_swb') %>%
  group_map(~ggplot(data = ., aes(x = original_value/10, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.7)+
              labs(x = 'Site water balance', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              #scale_x_continuous(trans = 'log10')+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))


p0 <- shapdata %>%
  group_by(feature) %>%
  filter(feature == 'OlsenP') %>%
  group_map(~ggplot(data = ., aes(x = original_value/10, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.7)+
              labs(x = 'Phosphorus availability (mg/kg)', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              coord_cartesian(ylim = c(-0.07,0.15))+
              scale_x_continuous(trans = 'log1p')+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))



p3<- shapdata %>%
  group_by(feature) %>%
  filter(feature == 'CHELSA_Annual_Precipitation') %>%
  group_map(~ggplot(data = ., aes(x = original_value, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.7)+
              labs(x = 'Annual precipitation (mm)', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              coord_cartesian(ylim = c(-0.07,0.15))+
              scale_x_continuous(trans = 'sqrt')+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))



p1<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'CHELSA_Min_Temperature_of_Coldest_Month') %>%
  group_map(~ggplot(data = ., aes(x = original_value/10, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.7)+
              labs(x = 'Min temperature of coldest month (\u00B0C)', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))


p2<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'CHELSA_Wind_max') %>%
  group_map(~ggplot(data = ., aes(x = original_value/1000, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.6)+
              labs(x = 'wind speed max （m/s）', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              scale_x_continuous(trans = 'sqrt')+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))


p5<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'Nitrogen') %>%
  group_map(~ggplot(data = ., aes(x = original_value, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.9)+
              labs(x = 'Soil nitrogen content (cg/kg)', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              coord_cartesian(ylim = c(-0.07,0.15))+
              scale_x_continuous(trans = 'log10')+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))


p7<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'SG_Silt_Content_0_100cm') %>%
  group_map(~ggplot(data = ., aes(x = original_value, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.7)+
              labs(x = 'Soil silt content (%)', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              coord_cartesian(ylim = c(-0.07,0.15))+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))
              #scale_x_continuous(trans = 'log10'))


p4<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'SG_Soil_pH_H2O_0_100cm') %>%
  group_map(~ggplot(data = ., aes(x = original_value/10, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
              geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.7)+
              labs(x = 'Soil pH', y = 'Contribution')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              coord_cartesian(ylim = c(-0.07,0.15))+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))
              #scale_x_continuous(trans = 'log10'))


p6<-shapdata %>%
  group_by(feature) %>%
  filter(feature == 'SG_Absolute_depth_to_bedrock') %>%
  group_map(~ggplot(data = ., aes(x = original_value/100, y = contribution))+
              geom_point(alpha = 0.1, color = '#1d3557', size  = 3)+
             geom_smooth(method = 'loess', color = '#000000', size = 1, se = TRUE, span = 0.7)+
              labs(x = 'Soil absolute depth to bedrock （cm）', y = 'Contribution')+
              scale_x_continuous(trans = 'sqrt')+
              geom_hline(yintercept = 0, linetype = 'dashed')+
              coord_cartesian(ylim = c(-0.07,0.15))+
              theme_classic()+
              theme(axis.title.x = element_text(size = 20, color = '#000000'), axis.title.y = element_blank(), axis.text = element_text(size = 20, color = '#000000')))
  

```


```{r}

library(ggpubr)

ggarrange(pp[[1]], p3[[1]], ppx[[1]],
          p7[[1]],p4[[1]], p5[[1]], 
          p0[[1]],ppy[[1]],p6[[1]],
            
          ncol = 3,
          nrow = 3,
          labels = c('A', 'B', 'C',
                     'D', 'E', 'F',
                     'G', 'H', 'I')
          )

```

