---
title: "Untitled"
output: html_document
date: "2024-04-08"
---

## This script is used for downloading individual occurrence data of those species that exhibit buttress roots gatherred from distinct studies in tropical regions. 


#### Load packages

```{r}
library(rgbif)
library(dplyr)

library(ggplot2)

library(readr)

library(feather)

library(sf)
library(rWCVP)

library(gbif.range)
library(terra)
library(rnaturalearth)
library(rWCVP)
library(rWCVPdata)


```




#### Load the data

#### The data is corrected from wcvp and deleted duplications on species. In total, 850 species are included in the buttress root species dataset. 

```{r}

correctdf <- read.csv('~/Desktop/buttress_root/buttress_species_correctnames.csv')

head(correctdf)

```

### Extract corrected names

```{r}

specieslist <- unique(correctdf$Accepted_SPNAME)


```


#### Start the loop to download.


```{r steup, include=FALSE}
#specieslist <- sub('^(\\S*\\s+\\S+).*', '\\1', specieslist)

for(i in 1:length(specieslist)){
  
  
  obs.pt <- get_gbif(sp_name = specieslist[i], basis = c('OBSERVATION', 'HUMAN_OBSERVATION'), time_period = c(1990,3000), duplicates = F, identic_xy = TRUE, absences = FALSE, grain = 990 )

## Several limitations should be used to filter speceis records. 

### 1) only observation and human observations are included
### 2) only recent data are included
### 3) relatively accurate spatial resolution
### 4) no duplication, no absence data

  obs.pt <- obs.pt %>%
    filter(grepl(specieslist[i], scientificName))


  obs.pt <- obs.pt %>%
  filter(taxonRank == 'SPECIES')
  if(nrow(obs.pt)>0){
  obs.pt$flag <- i
  
  nativerange <- try(wcvp_distribution(specieslist[i], taxon_rank = 'species', introduced = FALSE, extinct = FALSE, location_doubtful = FALSE), silent = TRUE)
  if("try-error" %in% class(nativerange)){
    obs.pt$native <- 'UNDET'
    write_csv(obs.pt, paste0('~/Desktop/buttress_root/gbif_species_dist/', i, '_observations_20240528.csv'))
  }else{

  obs <- obs.pt %>%
  st_as_sf(coords = c('decimalLongitude', 'decimalLatitude'), crs = st_crs(4326))
  obs$native <- st_intersects(obs, st_union(nativerange), sparse = FALSE)[,1]
  obs.pt$native <- obs$native

  
  write_csv(obs.pt, paste0('~/Desktop/buttress_root/gbif_species_dist/', i, '_observations_20240725.csv'))
  }
  }
}


```


```{r}
```

